Efficient inference, potential, and limitations of site-specific substitution models

### 2020-07-25 manuscript/response_to_referees.tex

\refb{Line 308: I fear this example analysis is quite (or too) technical and am not sure this will appeal to typical readers of Virus Evolution. Overall, the manuscript would be more suited for a more methodological journal like Systematic Biology or Bioinformatics.}

\response{We have made an effort to make this section more accessible, but we think it is an important part of the manuscript and one of the reasons we think Virus Evolution is an appropriate journal.
This section explores the ability to estimate site specific preferences from large and diverse alignments of RNA viruses.
Similar analysis could be done for other RNA viruses.
Furthermore, it points out implications for phylogenetic inferences.
We don't have solutions for the latter, but feel that it is still important to point them out.
}

\response{We have made an effort to simplify and clarify this example. But we feel that the application of to HIV data is one aspect of what makes this paper suitable for Virus Evolution: Diverse viruses like HIV are one place were data to apply such methods are available and the interpretation of conservation and viral fitness is of interest to many readers of Virus Evolution. }

\refb{But what I’m missing the most from the results and discussion sections of the manuscript, is clear advice on how to perform analyses on real data, what to look out for and how to potentially remedy the problem. The manuscript is too technical in that regard, and while assessing accuracy problems is important, a step-by-step approach for empirical data sets- to potentially/hopefully mitigate the problems mention in this manuscript - is even more important. It would be most helpful if the authors comment on the typical workflow, i.e. running ModelTest or something similar (PartitionFinder, automatic/included model selection in IQ-TREE, ...) and then sticking with the model found to infer the phylogeny. How often would one get in trouble with such an approach? And if problems arise, how severe are they? Could the clustering potentially be impacted or does the problem stay limited to certain branches (of certain length)? Or is the tree as a whole deemed older/younger than it really is, i.e. does the underestimation propagate through the tree all the way to the root or not?}

\response{
    Such inference from large viral data sets is increasingly feasible as data set size increases and complements deep mutations scanning approaches.
    While we identify problems for phylogenetic inference associated with strong variation model variation along the sequence, we don't have a solution.
}

@@

\Richard{See what model selectors do for simulated data}

\refb{In a way, I’m not too sure what the conclusions from Figure 3 are in practice? Everything except for the true model results in underestimation, even the site-specific model in TreeTime. From a practical point of view, what could be done to fix the underestimation problem? Would some form of post-hoc correction be advised, but how would one know which branches to correct and which to leave along? In that regard, could thorough knowledge of the data provide any guidance; but even then, and related to the HIV example (and other virus examples), inaccurate branch lengths of 100 years or more are most likely not a major concern.}

\response{In essence, Fig 3 shows that branch length underestimation caused by extensive variation of rates and preferences is very hard to avoid.
As outlined above, we don't really have a good fix for that, but we investigate the causes in Fig 4.
That said, the main result of the paper is that site-specific frequencies can be accurately estimated.
The inability is estimating the length of long branches in RNA viruses is a serious concern for dating zoonosis and exploring the history of viruses causing measles, dengue, lassa etc.
This problem is clearly apparent when trying to date the common ancestor of SIVs.
Molecular clock approaches suggest an ancestor a few hundred years in the past, while viruses in island populations show that the common ancestor is much more ancient.
}

@@

\response{
We are afraid there is a misunderstanding here.
This manuscript deals with GTR models and the root doesn't matter.
We only use root-to-tip distances in mid-point-rooted trees as a measure of divergence.
The only place where time comes in are the HIV examples, but we only use time as a proxy for evolutionary rate and don't do any molecular clock analysis.
All we want to do is to point out that branch length under-estimation potentially has consequences for dating rapidly evolving RNA viruses.}

\refb{Lines 651-653: I’m going to assume that most readers, like myself, will not be very familiar with the work of Halpern and Bruno (1998). Hence these sentences will need to provide more information regarding the following statement: ‘only allow for site-specific rates and equilibrium frequencies while using the same transition matrix for every site.’ Regarding the site-specific (evolutionary) rates, there may be confusion as to whether the process meant is actually among-site rate variation (typically modelled as a discretized gamma distribution according to various paper of Ziheng Yang), or relative rates between positions which can be superimposed on the former model. I’m going to assume the latter, but in that case it’s not clear whether for each site a unique rate will be estimated (if so, that’s a lot of additional parameters and this needs to be discussed) or whether classes of sites will be imposed (the three different codon positions for example) and that for only each of these classes a unique rate will be estimated? Essentially, I have the same question regarding the equilibrium frequencies, as it’s not clear from the methods section (but I can guess from the simulated data section; still, the authors should be more explicit here).}

\response{
The points raised by this reviewer have made it clear to us that the description of our basic motivation and model set-up was not clear enough. In response to this, we have rewritten several parts of the manuscript and point out explicitly in multiple places that we do not consider partition models but investigate to what extend models in which every site has their own set of of frequencies and an overall rate can be estimated. }

@@

\refb{Implementation: this type of modelling is not new, although it surely is a useful -- and needed -- extension to TreeTime. It’s unclear at this point whether the transition matrix will also be (potentially) site-specific or not, and I encourage the authors to mention this explicitly in the implementation paragraph.}

\response{We currently implement the model as specified in Equation 1. We now mention this explicitly and point out that site specific $W_{ij}$ are not currently implemented.}

\refb{Line 681: This makes it sounds like either two data partitions will be simulated that will either be concatenated or the contents will be scrambled to create one alignment of length L=2000 that is then a mixture of the different (site-specific) processes. But I have the feeling that this is not what the authors are after; can this be more explicitly explained?}

\response{The reviewer is correct, we don't have two data partitions of length 1000.
Every site has its own set of frequencies and an overall rate.
Instead, we produce hierarchical data sets that allow us to explore the effect of different aspects of the problem (tree, model realization, realization of the sequence evolution given the model) on the inferences.
We added sentence to make this clear. }

### 2020-07-25 manuscript/ms.tex
edits to discussion

\section*{Discussion}
Sequence evolution models attempt to capture some biological realism while being computationally tractable and easy to parameterize.
A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence or to allow for a small number of data partitions with different preferences.
Different positions in a DNA sequence or protein, however, differ markedly in the preferences for different states -- this site-specific preference is the basis of all common homology search tools.
While the need to account for this variation has been known for decades \citep{bruno1996modeling,halpern_evolutionary_1998}, it is typically not modeled due to over-fitting concerns and due to computational complexity.
However, amino acid preferences can now be measured with high-throughput \citep{fowler_deep_2014} and data set sizes are increasing rapidly such that more complex models can be estimated.
Large data sets, however, require efficient and scalable methods.

>>>>

\section*{Discussion}
Different positions in a genome sequence or a protein change at different rates and explore different subsets of the available states.
These site-specific properties are evolutionary conserved and are the basis of all common homology search tools.
Which states are permissible at which position can be measured with high-throughput using deep mutational scanning techniques \citep{fowler_deep_2014} or can be estimated from large alignments of homologous sequences.
Both approaches can reveal biological function and organization of the entity coded for by the sequence.
\citet{bruno1996modeling} pointed out that the frequencies of states in columns of an alignment are distorted by phylogenetic correlations and how these phylogenetic correlations can be accounted for.
However, estimating these frequencies accurately requires large data sets that have only recently become available.

While accurate estimates of site-specific frequencies require correction of phylogenetic correlations, accurate phylogenetic inference requires sufficiently realistic models of sequence evolution while being computationally tractable and easy to parameterize.
A common compromise is to model rate heterogeneity as random effects, while assuming identical site preferences across the sequence or to allow for a small number of data partitions with different preferences \citep{Lartillot01062004}.
